# (PART) DISTANCES MATRICES {-}

# Distance Calculation in Microbiome Data Analysis
Calculating distance matrices between samples using appropriate metrics is a fundamental step in data preprocessing. This practice efficiently captures the similarity or dissimilarity between samples, laying the groundwork for subsequent analysis and dimensionality reduction techniques such as PCA, MDS, or t-SNE.

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo  =TRUE,
  message  =FALSE,
  warning  =FALSE,
  cache  =FALSE,
  comment  =NA)
  
  library(tidyverse)

```


## Import processed and transformed data objects
```{r message=FALSE, warning=FALSE}
set.seed(110912)

library(phyloseq)
library(microbiome)
library(vegan)

load("data/ps_transformed.rda", verbose = TRUE)
load("data/phyloseq_objects.rda", verbose = TRUE)
load("data/phyloseq_extra_objects.rda", verbose = TRUE)
```



## Phyloseq Bray-Curtis Dissimilarity
The Bray-Curtis dissimilarity calculates the dissimilarity between two samples based on the relative abundances of taxa. It considers both presence and abundance information but is not sensitive to differences in abundance magnitude.

```{r message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)

ps_asin_bray_dist  <- vegdist(t(otu_table(ps_asin)), method="bray") %>%
  as.matrix()
ps_identity_bray_dist  <- vegdist(t(otu_table(ps_identity)), method="bray") %>%
  as.matrix 
ps_compositional_bray_dist  <- vegdist(t(otu_table(ps_compositional)), method="bray") %>%
  as.matrix 
ps_z_otu_bray_dist  <- vegdist(t(otu_table(ps_z_otu)), method="bray") %>%
  as.matrix 
ps_z_sample_bray_dist  <- vegdist(t(otu_table(ps_z_sample)), method="bray") %>%
  as.matrix 
ps_log10_bray_dist  <- vegdist(t(otu_table(ps_log10)), method="bray") %>%
  as.matrix 
ps_log10p_bray_dist  <- vegdist(t(otu_table(ps_log10p)), method="bray") %>%
  as.matrix 
ps_clr_bray_dist  <- vegdist(t(otu_table(ps_clr)), method="bray") %>%
  as.matrix 
ps_shift_bray_dist  <- vegdist(t(otu_table(ps_shift)), method="bray") %>%
  as.matrix 
ps_scale_bray_dist  <- vegdist(t(otu_table(ps_scale)), method="bray") %>%
  as.matrix 

save(ps_asin_bray_dist, 
     ps_compositional_bray_dist, 
     ps_z_otu_bray_dist, 
     ps_z_sample_bray_dist, 
     ps_log10_bray_dist, 
     ps_log10p_bray_dist, 
     ps_clr_bray_dist, 
     ps_shift_bray_dist, 
     ps_scale_bray_dist, 
     file = "data/bray_distances.rda")
```



## Phyloseq Extra Bray-Curtis Dissimilarity
In this section, we compute the Bray-Curtis dissimilarity distance from a phyloseq extra object. The Bray-Curtis dissimilarity distance is a measure of dissimilarity between microbial communities based on the abundance of different taxa present in each community. This analysis provides valuable insights into the compositional differences between microbial communities, which are essential for understanding microbial diversity and community structure in microbiome studies.

```{r message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)
library(microViz)

psextra_clr_asin_bray_dist  <- psextra_clr_dietswap %>% dist_calc(dist = "bray")
psextra_id_asin_bray_dist  <- psextra_id_dietswap %>% dist_calc(dist = "bray")
psextra_log10p_asin_bray_dist  <- psextra_log10p_dietswap %>% dist_calc(dist = "bray")

save(psextra_clr_asin_bray_dist, 
     psextra_id_asin_bray_dist, 
     psextra_log10p_asin_bray_dist, 
     file = "data/psextra_distances.rda")
```


## Confirm computed Bray-Curtis distance objects
```{r message=FALSE, warning=FALSE}
load("data/bray_distances.rda", verbose = T)
load("data/psextra_distances.rda", verbose = T)

```


## Jaccard Distance

The Jaccard distance measures the dissimilarity between two samples based on the presence or absence of taxa, ignoring their abundance information. It is calculated as the ratio of the number of taxa found in only one of the samples to the total number of taxa found in both samples.

```{r message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)

ps_asin_jaccard_dist  <- vegdist(t(otu_table(ps_asin)), method="jaccard") %>%
  as.matrix()
ps_identity_jaccard_dist  <- vegdist(t(otu_table(ps_identity)), method="jaccard") %>%
  as.matrix 
ps_compositional_jaccard_dist  <- vegdist(t(otu_table(ps_compositional)), method="jaccard") %>%
  as.matrix 
ps_z_otu_jaccard_dist  <- vegdist(t(otu_table(ps_z_otu)), method="jaccard") %>%
  as.matrix 
ps_z_sample_jaccard_dist  <- vegdist(t(otu_table(ps_z_sample)), method="jaccard") %>%
  as.matrix 
ps_log10_jaccard_dist  <- vegdist(t(otu_table(ps_log10)), method="jaccard") %>%
  as.matrix 
ps_log10p_jaccard_dist  <- vegdist(t(otu_table(ps_log10p)), method="jaccard") %>%
  as.matrix 
ps_clr_jaccard_dist  <- vegdist(t(otu_table(ps_clr)), method="jaccard") %>%
  as.matrix 
ps_shift_jaccard_dist  <- vegdist(t(otu_table(ps_shift)), method="jaccard") %>%
  as.matrix 
ps_scale_jaccard_dist  <- vegdist(t(otu_table(ps_scale)), method="jaccard") %>%
  as.matrix 

save(ps_asin_jaccard_dist, 
     ps_compositional_jaccard_dist, 
     ps_z_otu_jaccard_dist, 
     ps_z_sample_jaccard_dist, 
     ps_log10_jaccard_dist, 
     ps_log10p_jaccard_dist, 
     ps_clr_jaccard_dist, 
     ps_shift_jaccard_dist, 
     ps_scale_jaccard_dist, 
     file = "data/jaccard_distances.rda")
```

## Confirm computed Jaccard distance objects
```{r message=FALSE, warning=FALSE}
load("data/jaccard_distances.rda", verbose = T)

```


## UniFrac Distance (Weighted and Unweighted)
UniFrac distances quantify the dissimilarity between microbial communities based on the evolutionary distances along a phylogenetic tree. Unweighted UniFrac considers only the presence or absence of taxa, while weighted UniFrac also takes into account the relative abundance of taxa.


```{r message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)

ps_asin_jaccard_dist  <- vegdist(t(otu_table(ps_asin)), method="jaccard") %>%
  as.matrix()
ps_identity_jaccard_dist  <- vegdist(t(otu_table(ps_identity)), method="jaccard") %>%
  as.matrix 
ps_compositional_jaccard_dist  <- vegdist(t(otu_table(ps_compositional)), method="jaccard") %>%
  as.matrix 
ps_z_otu_jaccard_dist  <- vegdist(t(otu_table(ps_z_otu)), method="jaccard") %>%
  as.matrix 
ps_z_sample_jaccard_dist  <- vegdist(t(otu_table(ps_z_sample)), method="jaccard") %>%
  as.matrix 
ps_log10_jaccard_dist  <- vegdist(t(otu_table(ps_log10)), method="jaccard") %>%
  as.matrix 
ps_log10p_jaccard_dist  <- vegdist(t(otu_table(ps_log10p)), method="jaccard") %>%
  as.matrix 
ps_clr_jaccard_dist  <- vegdist(t(otu_table(ps_clr)), method="jaccard") %>%
  as.matrix 
ps_shift_jaccard_dist  <- vegdist(t(otu_table(ps_shift)), method="jaccard") %>%
  as.matrix 
ps_scale_jaccard_dist  <- vegdist(t(otu_table(ps_scale)), method="jaccard") %>%
  as.matrix 

save(ps_asin_jaccard_dist, 
     ps_compositional_jaccard_dist, 
     ps_z_otu_jaccard_dist, 
     ps_z_sample_jaccard_dist, 
     ps_log10_jaccard_dist, 
     ps_log10p_jaccard_dist, 
     ps_clr_jaccard_dist, 
     ps_shift_jaccard_dist, 
     ps_scale_jaccard_dist, 
     file = "data/jaccard_distances.rda")

```


## Euclidean Distance
The Euclidean distance calculates the straight-line distance between two samples in a multidimensional space based on their abundance profiles. It considers differences in both presence and abundance of taxa.

```{r message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)

ps_asin_euclidean_dist  <- vegdist(t(otu_table(ps_asin)), method="euclidean") %>%
  as.matrix()
ps_identity_euclidean_dist  <- vegdist(t(otu_table(ps_identity)), method="euclidean") %>%
  as.matrix 
ps_compositional_euclidean_dist  <- vegdist(t(otu_table(ps_compositional)), method="euclidean") %>%
  as.matrix 
ps_z_otu_euclidean_dist  <- vegdist(t(otu_table(ps_z_otu)), method="euclidean") %>%
  as.matrix 
ps_z_sample_euclidean_dist  <- vegdist(t(otu_table(ps_z_sample)), method="euclidean") %>%
  as.matrix 
ps_log10_euclidean_dist  <- vegdist(t(otu_table(ps_log10)), method="euclidean") %>%
  as.matrix 
ps_log10p_euclidean_dist  <- vegdist(t(otu_table(ps_log10p)), method="euclidean") %>%
  as.matrix 
ps_clr_euclidean_dist  <- vegdist(t(otu_table(ps_clr)), method="euclidean") %>%
  as.matrix 
ps_shift_euclidean_dist  <- vegdist(t(otu_table(ps_shift)), method="euclidean") %>%
  as.matrix 
ps_scale_euclidean_dist  <- vegdist(t(otu_table(ps_scale)), method="euclidean") %>%
  as.matrix 

save(ps_asin_euclidean_dist, 
     ps_compositional_euclidean_dist, 
     ps_z_otu_euclidean_dist, 
     ps_z_sample_euclidean_dist, 
     ps_log10_euclidean_dist, 
     ps_log10p_euclidean_dist, 
     ps_clr_euclidean_dist, 
     ps_shift_euclidean_dist, 
     ps_scale_euclidean_dist, 
     file = "data/euclidean_distances.rda")

```


## Aitchison Distance
The Aitchison distance, also known as the compositional distance or log-ratio distance, is specifically designed for compositional data such as microbiome abundance data. It calculates the dissimilarity between two samples based on the log-ratio transformations of their compositional vectors, providing meaningful comparisons between samples while addressing the compositional constraints inherent in relative abundance data.

- Computing aitchison using `microViz` packages

```{r message=FALSE, warning=FALSE}
library(microViz)
library(dplyr)

# ps_identity_aitchison_dists <- ps_asin %>% dist_calc("aitchison")
ps_GlobalPatterns_aitchison_dists <- ps_GlobalPatterns %>% dist_calc("aitchison")
ps_dietswap_aitchison_dists <- ps_dietswap %>% dist_calc("aitchison")
ps_kostic_crc_aitchison_dists <- ps_kostic_crc %>% dist_calc("aitchison")

save(ps_GlobalPatterns_aitchison_dists, 
     ps_dietswap_aitchison_dists, 
     ps_kostic_crc_aitchison_dists, 
     file = "data/aitchison_distances.rda")

```


